Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-19 doc/mc-paper.tex
more style changes

As can be seen, a POG is represents a partitioned-operation
formula with a sharing of common subformulas.  Indeed, every node in the graph can be viewed as a partitioned-operation formula, and so we write
$\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi_{\nodeu})$.

We define the {\em size} of POG $P$, written $|P|$ to be the
the number of nodes labeled $\pand$ or $\por$ plus the number of edges from these nodes to their children.  Ring
evaluation of $P$ can be performed with at most $|P|$ ring
operations by traversing the graph from the terminal nodes up to
the root, computing a value $\rep(\nodeu, w)$ for each node $\nodeu$.
The final result is then $\rep(\noder, w)$.

POGs generalize the d-DNNF graphs in two ways:
\begin{itemize}
\item They allow negation of arbitrary nodes in the graph, not just
  variables.
  %, and so we allow it in the interest of generality.
%%  Negation could be useful in some contexts, such as when
%%  converting binary decision diagrams (BDDs) having complemented
%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
\item They allow arbitrary arguments to a $\por$ operation, as long as
  they have disjoint models.  By contrast, each
  sum node $\nodeu$ in a d-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
%%Our generalization is provided to allow an
%%  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
%%  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  These generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

@@

Clauses can be added in two different ways.  One is when they serve as
the {\em defining clauses} for an extension variable.  This form
occurs only when defining $\pand$ and $\por$ operations, as is
described in Section~\ref{sect:crat}.  Clauses can also be added or
deleted based on {\em implication redundancy}.  That is, when clause
$C$ satisfies $\phi \imply C$ for formula $\phi$, then it can either
be added $\phi$ to create the formula $\phi \cup \{C\}$ or it can be deleted
from $\phi \cup \{C\}$ to create $\phi$.

We use {\em reverse unit propagation} (RUP) to certify
implication redundancy when adding or deleting
clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
RUP
is the core rule supported by standard
proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic. It provides a simple and efficient
way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
proved redundant with respect to formula $\phi$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
{\em antecedent} clauses, such that each $D_i$ is in $\phi$.
A RUP step
proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
that the combination of the antecedents plus the negation of $C$ leads
to a contradiction.  The negation of $C$ is the formula
$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
\overline{\lit}_p$ having a CNF representation consisting of $p$ unit
clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
check processes the clauses of the antecedent in sequence, inferring
additional unit clauses.  In processing clause $D_i$, if all but one
of the literals in the clause is the negation of one of the
accumulated unit clauses, then we can add this literal to the
accumulated set.  That is, all but this literal have been falsified,
and so it must be set to true for the clause to be satisfied.  The
final step with clause $D_k$ must cause a contradiction, i.e., all of
its literals are falsified by the accumulated unit clauses.

@@

The dependency sets for the arguments represented by each pair of
literals $\lit_i$
and $\lit_{j}$ must
be disjoint, for $1 \leq i < j \leq k$.  A product operation may have no arguments,
representing Boolean constant $1$.  The only clause added to the proof will be
the unit literal $v$.  A reference to literal $-v$ then provides a way
to representing constant $0$.

@@

The sequence of operator declarations, asserted clauses, and
clause deletions represents a systematic transformation of the input formula
into a POG formula.  Validating all of these steps serves to prove that the
POG is logically equivalent to the input formula.
At the completion of the proof, the following conditions must hold:
\begin{enumerate}
\item There is exactly one remaining clause that was added via RUP
  addition, and this is a unit clause consisting of root literal $r$.
\item All of the input clauses have been deleted.
\end{enumerate}
Let $\inputformula$ denote the input formula and $\pogformula$ represent the formula encoded by unit clause $r$ and the defining clauses for the POG nodes:
$\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$.  We can write $\modelset(P)$ as a shorthand for $\modelset(\pogformula)$, recognizing that
any assignment to the input variables implicitly defines the assignments to the extension variables.
The sequence of clause addition steps provide a {\em forward implication} proof that
$\modelset(\inputformula) \subseteq \modelset(P)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
assign $\assign(r) = 0$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
that
$\modelset(P) \subseteq \modelset(\inputformula)$.

@@

\section{Implementations}
We have implemented versions of the three programs that, along with the \dfour{} knowledge compiler,  form the
toolchain illustrated in Figure~\ref{fig:chain}.  One version,
including a proof generator, proof checker, and ring evaluator for
standard and weighted model counting, is designed to be fully
scalable.  The other version, consisting of a proof checker and
ring evaluator has been formally verified within the \lean{} proof
system.
%% Our long-term goal is to have a toolchain suitable for use in a
%% knowledge compiler competition, along the lines of the annual
%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
%% Our experiments
%% consider the benchmark problems and computing capabilities that would
%% arise in such a competition and evaluate whether our implementationed toolchain
%% could handle these problems.  As with the original proof-generating
%% SAT competitions, we assume that the proofs at a competition could be
%% checked with the unverified checker.  Having a verified checker
%% that works on some subset of the benchmarks would provide an
%% additional degree of certainty.

>>>>

In essence, a POG represents a partitioned-operation
formula with a sharing of common subformulas.  Indeed, every node in the graph can be viewed as a partitioned-operation formula, and so we write
$\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi_{\nodeu})$.

We define the {\em size} of POG $P$, written $|P|$, to be the
the number of nodes labeled $\pand$ or $\por$ plus the number of edges from these nodes to their children.  Ring
evaluation of $P$ can be performed with at most $|P|$ ring
operations by traversing the graph from the terminal nodes up to
the root, computing a value $\rep(\nodeu, w)$ for each node $\nodeu$.
The final result is then $\rep(\noder, w)$.

POGs generalize the d-DNNF graphs in two ways:
\begin{itemize}
\item They allow negation of arbitrary nodes in the graph, not just
  variables.
  %, and so we allow it in the interest of generality.
%%  Negation could be useful in some contexts, such as when
%%  converting binary decision diagrams (BDDs) having complemented
%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
\item They allow arbitrary arguments to a $\por$ operation, as long as
  these have disjoint models.  By contrast, each
  sum node $\nodeu$ in a d-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
%%Our generalization is provided to allow an
%%  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
%%  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  These generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

@@

Clauses can be added in two different ways.  One is when they serve as
the {\em defining clauses} for an extension variable.  This form
occurs only when defining $\pand$ and $\por$ operations, as is
described in Section~\ref{sect:crat}.  Clauses can also be added or
deleted based on {\em implication redundancy}.  That is, when clause
$C$ satisfies $\phi \imply C$ for formula $\phi$, then it can either
be added to $\phi$ to create the formula $\phi \cup \{C\}$ or it can be deleted
from $\phi \cup \{C\}$ to create $\phi$.

We use {\em reverse unit propagation} (RUP) to certify
implication redundancy when adding or deleting
clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
RUP
is the core rule supported by standard
proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic. It provides a simple and efficient
way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
proved redundant with respect to formula $\phi$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
{\em antecedent} clauses, such that each $D_i$ is in $\phi$.
A RUP step
proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
that the combination of the antecedents plus the negation of $C$ leads
to a contradiction.  The negation of $C$ is the formula
$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
\overline{\lit}_p$, having a CNF representation consisting of $p$ unit
clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
check processes the clauses of the antecedent in sequence, inferring
additional unit clauses.  In processing clause $D_i$, if all but one
of the literals in the clause is the negation of one of the
accumulated unit clauses, then we can add this literal to the
accumulated set.  That is, all but this literal have been falsified,
and so it must be set to true for the clause to be satisfied.  The
final step with clause $D_k$ must cause a contradiction, i.e., all of
its literals are falsified by the accumulated unit clauses.

@@

The dependency sets for the arguments represented by each pair of
literals $\lit_i$
and $\lit_{j}$ must
be disjoint, for $1 \leq i < j \leq k$.  A product operation may have no arguments,
representing Boolean constant $1$.  The only clause added to the proof will be
the unit literal $v$.  A reference to literal $-v$ then provides a way
of representing constant $0$.

@@

The sequence of operator declarations, asserted clauses, and
clause deletions represents a systematic transformation of the input formula
into a POG formula.  Validating all of these steps serves to prove that the
POG is logically equivalent to the input formula.
At the completion of the proof, the following conditions must hold:
\begin{enumerate}
\item There is exactly one remaining clause that was added via RUP
  addition, and this is a unit clause consisting of root literal $r$.
\item All of the input clauses have been deleted.
\end{enumerate}
In other words, at the end of the proof it must hold that the active clauses be exactly those
in $\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
of unit clause $r$ and the defining clauses of the POG nodes. We can write
$\modelset(P)$ as a shorthand for $\modelset(\pogformula)$, recognizing that
any assignment to the input variables implicitly defines the assignments to the extension variables.
Let $\inputformula$ denote the input formula.
The sequence of clause addition steps provides a {\em forward implication} proof that
$\modelset(\inputformula) \subseteq \modelset(P)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
assign $\assign(r) = 0$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
that
$\modelset(P) \subseteq \modelset(\inputformula)$.

@@

\section{Implementations}
We have implemented versions of the three programs that, along with the \dfour{} knowledge compiler,  form the
toolchain illustrated in Figure~\ref{fig:chain}.  One version,
including a proof generator, proof checker, and ring evaluator for
standard and weighted model counting, is designed to be fully
scalable.  The other version, consisting of a proof checker and
ring evaluator has been formally verified within the \lean{} theorem prover.
%% Our long-term goal is to have a toolchain suitable for use in a
%% knowledge compiler competition, along the lines of the annual
%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
%% Our experiments
%% consider the benchmark problems and computing capabilities that would
%% arise in such a competition and evaluate whether our implementationed toolchain
%% could handle these problems.  As with the original proof-generating
%% SAT competitions, we assume that the proofs at a competition could be
%% checked with the unverified checker.  Having a verified checker
%% that works on some subset of the benchmarks would provide an
%% additional degree of certainty.
